---
title: "Lab 18- Pertussis mini-project"
author: "Emily Ignatoff (A16732102)"
format: pdf
---

Pertussis (aka Whooping Cough) is a deadly lung infection caused by the bacteria *Bordatella pertussis*.

The CDC tracks pertussis cases around the US: https://tinyurl.com/pertussiscdc

We can "scrape" this data using the R **datapasta** package.

```{r, include=FALSE}
cdc <- data.frame(
  year = c(1922L,1923L,1924L,1925L,
           1926L,1927L,1928L,1929L,1930L,1931L,
           1932L,1933L,1934L,1935L,1936L,
           1937L,1938L,1939L,1940L,1941L,1942L,
           1943L,1944L,1945L,1946L,1947L,
           1948L,1949L,1950L,1951L,1952L,
           1953L,1954L,1955L,1956L,1957L,1958L,
           1959L,1960L,1961L,1962L,1963L,
           1964L,1965L,1966L,1967L,1968L,1969L,
           1970L,1971L,1972L,1973L,1974L,
           1975L,1976L,1977L,1978L,1979L,1980L,
           1981L,1982L,1983L,1984L,1985L,
           1986L,1987L,1988L,1989L,1990L,
           1991L,1992L,1993L,1994L,1995L,1996L,
           1997L,1998L,1999L,2000L,2001L,
           2002L,2003L,2004L,2005L,2006L,2007L,
           2008L,2009L,2010L,2011L,2012L,
           2013L,2014L,2015L,2016L,2017L,2018L,
           2019L,2020L,2021L,2022L,2024L),
  cases = c(107473,164191,165418,152003,
            202210,181411,161799,197371,
            166914,172559,215343,179135,265269,
            180518,147237,214652,227319,103188,
            183866,222202,191383,191890,109873,
            133792,109860,156517,74715,69479,
            120718,68687,45030,37129,60886,
            62786,31732,28295,32148,40005,
            14809,11468,17749,17135,13005,6799,
            7717,9718,4810,3285,4249,3036,
            3287,1759,2402,1738,1010,2177,2063,
            1623,1730,1248,1895,2463,2276,
            3589,4195,2823,3450,4157,4570,
            2719,4083,6586,4617,5137,7796,6564,
            7405,7298,7867,7580,9771,11647,
            25827,25616,15632,10454,13278,
            16858,27550,18719,48277,28639,32971,
            20762,17972,18975,15609,18617,6124,2116,3044,35493)
)
```

```{r}
head(cdc)
```


> Q1 Let's make a ggplot with these data:

```{r}
library(ggplot2)
ggplot(cdc) + aes(x=year, y=cases) + 
  geom_line()
```

>Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

```{r}
ggplot(cdc) + aes(x=year, y=cases) + 
  geom_line() + geom_vline(xintercept=1946, color="purple", lwd=1) +
  geom_vline(xintercept=1996, color= "turquoise2", lwd=1) +
  geom_vline(xintercept=2020, color= "red", lwd=1) +
  geom_vline(xintercept=2004, color="green3", lwd=1)
```

There were high case numbers before the first wP (whole cell) vaccine in 1946. Case number dropped drastically in the years following vaccine introduction to a low in the 1970s. After the introduction of the aP vaccine, around 2004, cases begin to peak again. Cases dip during the COVID pandemic, and rise rapidly following it.

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

After the introduction of the aP vaccine, cases start to trend upwards again (but nowhere near as high as before). Various factors can contribute to this, such as immunity waning without a booster shot, varying vaccination rates, and varying exposure to other people such as during quarantine. Perhaps the aP vaccine interacts with the immune system differently than the wP vaccine.

Q. What is different about the immune response to infection if you had an older wP vaccine vs the newer aP vaccine?

##CMI-PB (Computational Models of Immunity: Pertussis Boost)

The CMI-PB project aims to address this key question: what is different between aP and wP individuals?

We can get all the data from this project via JSON API calls
To do this we will install the **jsonlite** package
```{r}
library(jsonlite)
subject <- read_json("https://www.cmi-pb.org/api/v5_1/subject", 
                     simplifyVector = TRUE)

head(subject)

```

>Q How many individuals (subjects) are in this dataset?

```{r}
nrow(subject)
```


>Q. How many wP and aP primed individuals are in this dataset?

```{r}
table(subject$infancy_vac)
```


>Q. How many male and female subjects are in the study?

```{r}
table(subject$biological_sex)
```

```{r}
table(subject$race, subject$biological_sex)
```

This is not representative of the US population 

Let's get the rest of the data we want:

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen",
                       simplifyVector = T)
antibody <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer",
                      simplifyVector = T)
```

```{r}
head(specimen)
```

```{r}
head(antibody)
```

I now have three tables of data from the CMI-PB project: `subject`, `specimen`, and `antibody`.

How can we put all these back together? We can use the **dplyr** functions `*_join()`
In our case we will use an `inner_join()` for the data to avoid keeping individuals for whom follow-up testing was not possible.

```{r}
library(dplyr)
meta <- inner_join(subject, specimen)
head(meta)
```

```{r}
dim(subject)
dim(specimen)
dim(meta)
```

Now we can join the `antibody` data to the `meta` table we just generated:

```{r}
abdata <- inner_join(meta, antibody)
head(abdata)
```

>Q. How many different antibody isotypes are there in this dataset?

```{r}
length(abdata$isotype)
```


```{r}
table(abdata$isotype)
```
```{r}
table(abdata$antigen)
```

I want a plot of antigen levels across the whole dataset.

```{r}
ggplot(abdata) + aes(x=MFI, y=antigen) +
  geom_boxplot()
```

```{r}
ggplot(abdata) + aes(x=MFI_normalised, y=antigen) +
  geom_boxplot()
```

Antigens like FIM2/3, PT, FELD1 have quite a large range of values. Others like measles don't show much activity.


>Q. Are there differences at this whole-dataset level between aP and wP vaccinated individuals?

```{r}
ggplot(abdata) + aes(x=MFI_normalised, y=antigen, col=infancy_vac) +
  geom_boxplot()
```

```{r}
ggplot(abdata) + aes(x=MFI_normalised, y=antigen, col=infancy_vac) +
  geom_boxplot() + facet_wrap(~infancy_vac)
```

## Examine IgG Ab titer levels

For this we need to isolate just isotype IgG.

```{r}
igg <- abdata %>% filter(isotype == "IgG")
head(igg)
```

An overview boxplot:

```{r}
ggplot(igg) + aes(x=MFI_normalised, y=antigen, col=infancy_vac) +
  geom_boxplot()
```

Digging further to look at the time course of IgG isotype PT antigen levels across aP and wP individuals


```{r}
#filter to only include 2021 data
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

#filter to look at igg and pt only
pt.igg <- abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT")

#plot 2021 data, colored by infancy vac (aP vs wP)
ggplot(pt.igg) +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) +
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)")
```




