---
title: "Comparative Structure Analysis"
author: "Emily Ignatoff (A16732102)"
format: html
---

We will use a "pipeline" to analyze the structure of Adenylate Kinase. This begins we a single sequence identifier and ends in a PCA.

```{r}
library(bio3d)
```

Step 1. Retrieve a sequence for the protein we are interested in. We'll take ADK (1ake_A) for our protein:

```{r}
id <- "1ake_A"
aa <- get.seq(id)
aa
```


Step 2. Run a BLAST search of the PDB for all related sequences to our ADK sequence
```{r}
blast <- blast.pdb(aa)
```

```{r}
hits <- plot(blast)
```

These are my "top hits" i.e. all the structures in the PDB database that are related to our input sequence.

```{r}
hits$pdb.id
```

Step 3. Download structures
```{r}
files <- get.pdb(hits$pdb.id, path="pdbs", split=TRUE, gzip=TRUE)
```

Step 4. Align and Superimpose

```{r}
pdbs <- pdbaln(files, fit = TRUE, exefile="msa")
```

```{r}
pdbs
```

Step 5. run a PCA

Let's use a PCA to make sense of how these strucutres all relate to each other.

```{r}
pc <- pca(pdbs)
plot(pc)
```


Let's make a trajectory (or movie) of the main conformational changes by PC1. We will use `mktrj()`

```{r}
mktrj(pc, file="pca_result.pdb")
```


Back of the envelope comparison of the PDB size to Uniprot:

```{r}
pdb <- 231029
uniprot <- 253206171

pdb/uniprot * 100
```

