---
title: "Class 12- Genome Informatics"
author: "Emily Ignatoff (A16732102)"
format: pdf
---


> Q5. What proportion of the Mexican Ancestry in Los Angeles sample population (MXL) are
homozygous for the asthma associated SNP (G|G)?

```{r}
mxl <- read.csv('mexican_la.csv')

table(mxl$Genotype..forward.strand.)
```
# Homework: Population Scale Analysis

I will be looking at 230 samples of RNASeq for the asthma associated SNPs

```{r}
data1 <- read.csv("popdata.csv")
```


>Q13. How many of each genotype and what are their corresponding median expression levels?

```{r}
table(data1$geno)
```

```{r}
median_by_geno <- tapply(data1$exp, data1$geno, median)
print(median_by_geno)

```

>Q14. Generate a boxplot with a box per genotype, what could you infer from the relative
expression value between A/A and G/G displayed in this plot? Does the SNP effect the
expression of ORMDL3?

```{r}
library(ggplot2)
by_geno <- ggplot(data1) + aes(x=geno, y=exp, fill=geno) + geom_boxplot() +
  labs(xlab="Genotype", ylab="Expression level", title="Expression level by Genotype")
by_geno
```

Expression of A|A is much higher than expression of G|G. This SNP likely reduces the expression of ORDML3.